---
title: "IBMs"
author: "Bell the best"
date: "2023-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Individual based models eg Agent-based models

Models systems of discrete individuals and simulate changes in indivudal composition and properties. Can be used in conjunction with experimental data to predict dynamics of complex systems. 

*Perks*:(1) Includes among individual variation x. genotype, phenotype or location. 
        (2)Indivduals are discrete, so stochasticty is naturally apart of the model. x. genetic drift.
        
Individuals are represented in a list, table, matrix etc. 

### For example:

```{r}
inds <- array(data = 0, dim =c (5,3))
colnames(inds) <- paste0("Characteristic ", 1:ncol(inds))
rownames(inds) <- paste0("Individual ", 1:nrow(inds))
print(inds)
```

## Characteristics

Choosing which characteristics to model is of utmost important. It can be anything from phenotype, to location, to temporary statuses, like breeding. Whatever characteristics we choose, these will the biological process that we're trying to understand with out model.

So lets give our individuals some characteristics.

```{r}
colnames(inds) <- c("Body Mass", "x location", "y location")

```

Body mass being normally distributed is a logical assumption. Roe deer roughly has a mean around 23 with standard deviation of 3. So lets use that as an example. Though it does not necessarily have to be roe deer in mind.

```{r}
inds[, 1] <- rnorm(n = nrow(inds), mean = 23, sd = 3)   #writing inds[,1] is the same as inds[1:5, 1]

dim(inds)  #returns dimensions of matrix or array as a list [1] rows and [2] columns

```

Lets also give each individual a location in a 8x8 grid. Which will allow us to model distance and visually present information regarding relative position. 

```{r}
inds[, 2] <- sample(x = 1:8, size = nrow(inds), replace = T)
inds[, 3] <- sample(x=1:8, size = nrow(inds), replace = T)
```

This enables us to plot them in our imaginary landscape:

```{r}
plot(
  x = inds[, 2],              # x-coordinate data
  y = inds[, 3],              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  cex = 2,                    # Point size (enlarged by a factor of 2)
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
)

```

## Movement of individuals

Lets allow individuals to increase to decrease their x and y location by 1 cell. 

```{r}

x_move <- sample(x = c(-1, 0, 1), size = nrow(inds), replace = T)
y_move <- sample(x = c(-1, 0, 1), size = nrow(inds), replace = T)

inds[, 2] <- inds[, 2] + x_move
inds[, 3] <- inds[, 3] + y_move
plot(
  x = inds[, 2] - x_move,              # x-coordinate data
  y = inds[, 3] - y_move,              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  cex = 2,                    # Point size (enlarged by a factor of 2)
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
)

points(x = inds[, 2],              # x-coordinate data
  y = inds[, 3],              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  cex = 2,                    # Point size (enlarged by a factor of 2)
  col = ("mediumpurple"),
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
  )
arrows(
  x0 = inds[, 2] - x_move,  # Starting x-coordinates (old positions)
  y0 = inds[, 3] - y_move,  # Starting y-coordinates (old positions)
  x1 = inds[, 2],          # Ending x-coordinates (new positions)
  y1 = inds[, 3],          # Ending y-coordinates (new positions)
  length = 0.1,            # Length of the arrowheads
  angle = 30,              # Angle of the arrowheads
  code = 2,                # Type of arrowhead (2 for closed arrowheads)
  col = "grey",            # Arrow color
  lwd = 2                  # Modifies the thickness of the arrow
)
```

Although the above code is fine it is a good idea to create the same effect using a function:

```{r}
movement <- function(inds, xloc = 2, yloc = 3){
    total_inds   <- dim(inds)[1] # Get the number of individuals in inds by checking number of rows
    move_dists   <- c(-1, 0, 1)  # Define the possible distances to move
    x_move       <- sample(x = move_dists, size = total_inds, replace = TRUE)
    y_move       <- sample(x = move_dists, size = total_inds, replace = TRUE)
    inds[, xloc] <- inds[, xloc] + x_move
    inds[, yloc] <- inds[, yloc] + y_move
    return(inds)
}
```
 
We can now easily move indivduals by using this line of code:

```{r}
inds <- movement(inds)
```

Lets have our roe deer run around for a while in a simulation:

```{r}
time_steps <- 20
for (i in 1:time_steps)
  inds <- movement(inds)

```

But as we had a boundary of 8x8 some of our individuals have run away, we can't have that so lets put up a digital bouncy fence.

```{r}
movement <- function(inds, xloc = 2, yloc = 3, xmax = 8, ymax = 8){
    total_inds   <- dim(inds)[1]; # Get the number of individuals in inds
    move_dists   <- c(-1, 0, 1);  # Define the possible distances to move
    x_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    y_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    inds[, xloc] <- inds[, xloc] + x_move;
    inds[, yloc] <- inds[, yloc] + y_move;
    # =========   The reflecting boundary is added below
    for(i in 1:total_inds){               # For each individual i in the array
        if(inds[i, xloc] > xmax){         # If it moved passed the maximum xloc
            inds[i, xloc] <- xmax - 1;    # Then move it back toward the centre
        }
        if(inds[i, xloc] < 1){            # If it moved below 1 on xloc
            inds[i, xloc] <- 2;           # Move it toward the centre (2)
        }
        if(inds[i, yloc] > ymax){         # If it moved passed the maximum yloc
            inds[i, yloc] <- ymax - 1;    # Then move it back toward the centre
        }
        if(inds[i, yloc] < 1){            # If it moved below 1 on yloc
            inds[i, yloc] <- 2;           # Then move it toward the centre (2)
        }
    } 
    # =========  Now all individuals should stay on the landscape
    return(inds);
}

inds           <- array(data = 0, dim = c(5, 3))
colnames(inds) <- c("body_mass", "x_loc", "y_loc")
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)

plot(
  x = inds[, 2] - x_move,              # x-coordinate data
  y = inds[, 3] - y_move,              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  cex = 2,                    # Point size (enlarged by a factor of 2)
  col = rgb(1, 0.5, 0.8),
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
)

time_steps <- 20

for (i in 1:time_steps)
  inds <- movement(inds)
  points(x = inds[i, 2],              # x-coordinate data
         y = inds[i, 3],              # y-coordinate data
         pch = 16,                   # Point character (solid circle)
         cex = 2,                    # Point size (enlarged by a factor of 2)
         col = rgb((1/i), 0.5, 0.8, alpha = 0.7),
  )


```


