---
title: "IBMs"
author: "Bell the best"
date: "2023-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*(This code was made using https://bradduthie.github.io/blog/individual-based-models-in-r/ guide.)*

## Individual based models eg Agent-based models

Models systems of discrete individuals and simulate changes in indivudal composition and properties. Can be used in conjunction with experimental data to predict dynamics of complex systems. 

*Perks*:(1) Includes among individual variation x. genotype, phenotype or location. 
        (2)Indivduals are discrete, so stochasticty is naturally apart of the model. x. genetic drift.
        
Individuals are represented in a list, table, matrix etc. 


### For example:

```{r}
inds <- array(data = 0, dim =c (5,3))
colnames(inds) <- paste0("Characteristic ", 1:ncol(inds))
rownames(inds) <- paste0("Individual ", 1:nrow(inds))
print(inds)
```

## Characteristics

Choosing which characteristics to model is of utmost important. It can be anything from phenotype, to location, to temporary statuses, like breeding. Whatever characteristics we choose, these will the biological process that we're trying to understand with out model.

So lets give our individuals some characteristics.

```{r}
colnames(inds) <- c("Body Mass", "x location", "y location")

```

Body mass being normally distributed is a logical assumption. Roe deer roughly has a mean around 23 with standard deviation of 3. So lets use that as an example. Though it does not necessarily have to be roe deer in mind.

```{r}
inds[, 1] <- rnorm(n = nrow(inds), mean = 23, sd = 3)   #writing inds[,1] is the same as inds[1:5, 1]

dim(inds)  #returns dimensions of matrix or array as a list [1] rows and [2] columns

```

Lets also give each individual a location in a 8x8 grid. Which will allow us to model distance and visually present information regarding relative position. 

```{r}
inds[, 2] <- sample(x = 1:8, size = nrow(inds), replace = T)
inds[, 3] <- sample(x=1:8, size = nrow(inds), replace = T)
```

This enables us to plot them in our imaginary landscape:

```{r}
plot(
  x = inds[, 2],              # x-coordinate data
  y = inds[, 3],              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  cex = 2,                    # Point size (enlarged by a factor of 2)
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
)

```

## Movement of individuals

Lets allow individuals to increase to decrease their x and y location by 1 cell. 

```{r}

x_move <- sample(x = c(-1, 0, 1), size = nrow(inds), replace = T)
y_move <- sample(x = c(-1, 0, 1), size = nrow(inds), replace = T)

inds[, 2] <- inds[, 2] + x_move
inds[, 3] <- inds[, 3] + y_move
plot(
  x = inds[, 2] - x_move,              # x-coordinate data
  y = inds[, 3] - y_move,              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  cex = 2,                    # Point size (enlarged by a factor of 2)
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
)

points(x = inds[, 2],              # x-coordinate data
  y = inds[, 3],              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  cex = 2,                    # Point size (enlarged by a factor of 2)
  col = ("mediumpurple"),
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
  )
arrows(
  x0 = inds[, 2] - x_move,  # Starting x-coordinates (old positions)
  y0 = inds[, 3] - y_move,  # Starting y-coordinates (old positions)
  x1 = inds[, 2],          # Ending x-coordinates (new positions)
  y1 = inds[, 3],          # Ending y-coordinates (new positions)
  length = 0.1,            # Length of the arrowheads
  angle = 30,              # Angle of the arrowheads
  code = 2,                # Type of arrowhead (2 for closed arrowheads)
  col = "grey",            # Arrow color
  lwd = 2                  # Modifies the thickness of the arrow
)
```

Although the above code is fine it is a good idea to create the same effect using a function:

```{r}
movement <- function(inds, xloc = 2, yloc = 3){
    total_inds   <- dim(inds)[1] # Get the number of individuals in inds by checking number of rows
    move_dists   <- c(-1, 0, 1)  # Define the possible distances to move
    x_move       <- sample(x = move_dists, size = total_inds, replace = TRUE)
    y_move       <- sample(x = move_dists, size = total_inds, replace = TRUE)
    inds[, xloc] <- inds[, xloc] + x_move
    inds[, yloc] <- inds[, yloc] + y_move
    return(inds)
}
```
 
We can now easily move indivduals by using this line of code:

```{r}
inds <- movement(inds)
```

Lets have our roe deer run around for a while in a simulation:

```{r}
time_steps <- 20
for (i in 1:time_steps)
  inds <- movement(inds)

```

But as we had a boundary of 8x8 some of our individuals have run away, we can't have that so lets put up a digital bouncy fence.

```{r}
movement <- function(inds, xloc = 2, yloc = 3, xmax = 8, ymax = 8){
    total_inds   <- dim(inds)[1]; # Get the number of individuals in inds
    move_dists   <- c(-1, 0, 1);  # Define the possible distances to move
    x_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    y_move       <- sample(x = move_dists, size = total_inds, replace = TRUE);
    inds[, xloc] <- inds[, xloc] + x_move;
    inds[, yloc] <- inds[, yloc] + y_move;
    # =========   The reflecting boundary is added below
    for(i in 1:total_inds){               # For each individual i in the array
        if(inds[i, xloc] > xmax){         # If it moved passed the maximum xloc
            inds[i, xloc] <- xmax - 1;    # Then move it back toward the centre
        }
        if(inds[i, xloc] < 1){            # If it moved below 1 on xloc
            inds[i, xloc] <- 2;           # Move it toward the centre (2)
        }
        if(inds[i, yloc] > ymax){         # If it moved passed the maximum yloc
            inds[i, yloc] <- ymax - 1;    # Then move it back toward the centre
        }
        if(inds[i, yloc] < 1){            # If it moved below 1 on yloc
            inds[i, yloc] <- 2;           # Then move it toward the centre (2)
        }
    } 
    # =========  Now all individuals should stay on the landscape
    return(inds);
}

inds           <- array(data = 0, dim = c(5, 3))
colnames(inds) <- c("body_mass", "x_loc", "y_loc")
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)


time_steps <- 20

for (i in 1:time_steps){
  inds <- movement(inds)
  }


```

To save each time step we have to adjust it slightly:

```{r}

time_steps <- 20
inds_hist  <- NULL; # Here's the list
for(i in 1:time_steps){
    inds            <- movement(inds)
    inds_hist[[i]] <- inds; # Add to list
}
print(inds)

for (i in 1:5) {
  print(paste0("Year ", i))
  print(inds_hist[[1]])
} 

```

We can then extract the information for indivdual 1 over the 20 years:

```{r}

indi_1 <- matrix(data= NA, nrow=20, ncol = 3)
colnames(indi_1) <- c("Year", "x-location", "y-location")

for(i in 1:time_steps){
  indi_1[i, 1] <- i    #saves timestep in first column
  indi_1[i, 2] <- inds_hist[[i]][1,2]  #extracts x(2) value for the i timestep of individual 1 (1)
  indi_1[i, 3] <- inds_hist[[i]][1,3]  # -'- but for the y(3) value
}

print(indi_1)

colorrange <- colorRampPalette(c("chartreuse", "cyan"))
my.colorrange <- colorrange(20)

plot(
  x = indi_1[, 2],              # x-coordinate data
  y = indi_1[, 3],              # y-coordinate data
  pch = 16,                   # Point character (solid circle)
  col = my.colorrange,
  cex = 2,                    # Point size (enlarged by a factor of 2)
  xlim = c(1, 8),             # X-axis limits (range from 1 to 8)
  ylim = c(1, 8),             # Y-axis limits (range from 1 to 8)
  xlab = "x location",        # X-axis label
  mar = c(5, 5, 1, 1),        # Margins for the plot (bottom, left, top, right)
  ylab = "y location",        # Y-axis label
  cex.lab = 1.5,              # Size of axis labels (enlarged by a factor of 1.5)
  cex.axis = 1.5              # Size of axis values (enlarged by a factor of 1.5)
)

```
Adding offspring and birth to our model is the next step:


```{r}
inds           <- array(data = 0, dim = c(5, 4))
colnames(inds) <- c("body_mass", "x_loc", "y_loc", "repr")
rownames(inds) <- c("ind_1", "ind_2", "ind_3", "ind_4", "ind_5")
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = dim(inds)[1], replace = TRUE)

print(inds)

birth <- function(inds, lambda = 0.5, repr_col = 4) {
  total.inds <- nrow(inds)
  inds.col <- ncol(inds)
  inds[, repr_col] <- rpois(n =total.inds, lambda = lambda)
  total.offspring <- sum(inds[,repr_col])
      # the above code is reproduction, the below code is making new rows for the offspring. 
  new.inds <- matrix(data = 0, nrow = total.offspring, ncol = inds.col)
  new.inds[, 1] <- rnorm(n = nrow(new.inds), mean = 23, sd = 3)
  new.inds[, 2] <- sample(x = 1:8, size = nrow(new.inds), replace = T)
  new.inds[, 3] <- sample(x = 1:8, size = nrow(new.inds), replace = T)
      # Adding the offspring to the adults
  inds <- rbind(inds, new.inds)
  return(inds)
}

#New generation
inds <- birth(inds = inds)
rownames(inds) <- paste0("Individual ", 1:nrow(inds))
print(inds)


```
We can then simulate both movement and birth:

```{r}
time.steps <- 10
inds.hist <- NULL

for( i in 1:time.steps){
  inds <- movement(inds)
  inds <- birth(inds)
  inds.hist[[i]] <- inds
}

#Number of indivduals after 10 time steps:
print(nrow(inds.hist[[10]]))

ind.abund <- matrix(data = NA, nrow = 10, ncol = 2)

for (i in 1:time.steps) {
  ind.abund[i,1] <- i
  ind.abund[i,2] <- nrow(inds.hist[[i]])
}

print(ind.abund)


```

This model will grow exponentially, since there is no death and no density dependance. So lets also simulate some density dependent death. Since we have a spatial model lets assume each cell can only hold one indivdual. 


```{r}
inds <- matrix(data = 0, nrow = 5, ncol = 5)
colnames(inds) <- c("body.mass", "loc.x", "loc.y", "reproduction", "death") 
rownames(inds) <- paste0("Individual ", 1:nrow(inds))

inds[,1]       <- rnorm(n = nrow(inds), mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = nrow(inds), replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = nrow(inds), replace = TRUE)

death <- function(inds, xlen = 8, ylen = 8, xcol = 2, ycol = 3, dcol = 5){
  for (xdim in 1:xlen){
    for(ydim in 1:ylen){
      on.cell <- sum(inds[,xcol] == xdim & inds[, ycol] == ydim)
      if (on.cell > 1) {
        occupants <- which(inds[, xcol] == xdim & inds[, ycol] == ydim)   #occupants who live on the overpopulated cell, i.e. their row number
        random <- sample(x = occupants, size = on.cell - 1)    # randomly chooses all but on of the individuals from the vector on the cell 
        inds[random, dcol] <- 1             # sets individuals sampled death to 1
      }
    }
  }
  return(inds)
}

inds[2, 2] <- inds[1, 2]; # Individiual 2 now in same x location as 1
inds[2, 3] <- inds[1, 3]; # Individiual 2 now in same x location as 1
inds[3, 2] <- inds[1, 2]; # Individiual 3 now in same x location as 1
inds[3, 3] <- inds[1, 3]; # Individiual 3 now in same x location as 1

inds <- death(inds = inds)

inds <- inds[inds[, 5] == 0, ] # Removes all individuals who do not have a 0 in their fifth column, i.e. removes dead individuals. the , is important, dont forget it.  
print(inds)

```

Lets put it all together :) (movement, birth and death):

```{r}
# Create individuals

inds <- matrix(data = 0, nrow = 5, ncol = 5)
colnames(inds) <- c("body.mass", "loc.x", "loc.y", "reproduction", "death") 
rownames(inds) <- paste0("Individual ", 1:nrow(inds))

inds[,1]       <- rnorm(n = nrow(inds), mean = 23, sd = 3)
inds[,2]       <- sample(x = 1:8, size = nrow(inds), replace = TRUE)
inds[,3]       <- sample(x = 1:8, size = nrow(inds), replace = TRUE)

time.steps <- 40
inds.hist <- NULL
for (i in 1:time.steps){
  inds <- movement(inds)
  inds <- birth(inds)
  inds <- death(inds)
  inds <- inds[inds[, 5] == 0, ]
  inds.hist[[i]] <- inds
}

ind.abund <- matrix(data = NA, nrow = time.steps, ncol = 2)
for (i in 1:time.steps){
  ind.abund[i, 1] <- i
  ind.abund[i, 2] <- nrow(inds.hist[[i]])
}

plot(x = ind.abund[,1], y =ind.abund[,2], ylab = "Number of Individuals", xlab = "Years", col = "darkslategrey", pch = 16)

```

## Predator prey dynamics

Using the previous code inds to represent prey we will now add predators in a similar way:

```{r}
pred <-  matrix(data = 0, nrow = 5, ncol = 5)
colnames(pred) <- c("bodymass", "x location", "y location", "Birth", "Death")
rownames(pred) <- paste0("Predator ", 1:nrow(pred))

pred[, 1] <- rnorm(n = nrow(pred), mean = 20, sd = 2.5)  #This approximates a lynx ish
pred[, 2] <- sample(x = 1:8, size = nrow(pred), replace = T)
pred[, 3] <- sample(x = 1:8, size = nrow(pred), replace = T)

```

We can assume predators only move 1 cell for simplicity, meaning they can use the same function as prey. But their birth and death should be based upon whether or not they've managed to eat.

```{r}

predation <- function(pred, inds, xcol = 2, ycol = 3, rcol = 4, dcol = 5){
  predators <- nrow(pred)  # is the number of predators
  pred[, dcol] <- 1         #dead unless something changes
  pred[, rcol] <- 0         #No baby unless something changes
  #something changes-----------------------
  for (p in 1:predators){
    xloc <- pred[p, xcol]  #Extracts location of predator p
    yloc <- pred[p, ycol]
    N.prey <- sum(inds[, xcol] == xloc & inds[, ycol] == yloc)  #sums all the prey on p's location
    #predation starts
    if(N.prey > 0){
      pred[p, dcol] <- 0   #The predator set to not die
    }
    if(N.prey > 1){
      pred[p, rcol] <- 1   #Predator set reproduces
    }
    # Unfortunately something also happens to the prey: -------------------
    if(N.prey > 0) {
      prey <- which(inds[, xcol] == xloc & inds[, ycol] == yloc)   #looks for all prey were predator p and prey coexist
      if(N.prey > 2){
        prey <- sample(x = prey, size = 2, replace = FALSE)        #Makes it so that if if there are more than two prey on the cell only two get eaten. Samples two from among the prey.
      }
      inds[prey, dcol] <- 1  #Records prey as dead.
    }
  }  #This for loop recorded which predators dies, survive or reproduce, and which prey gets predated upon.
  # Remove predators that have died ---------------------------
  pred <- pred[pred[,dcol] == 0,]
  # Add the offspring into predator pool
  pred.off <- sum(pred[, rcol])
  new.pred <- matrix(data = 0, nrow = pred.off, ncol = ncol(pred))
  new.pred[, 1] <- rnorm(n = nrow(new.pred), mean = 20, sd = 2.5)  #This approximates a lynx ish
  new.pred[, 2] <- sample(x = 1:8, size = nrow(new.pred), replace = T)
  new.pred[, 3] <- sample(x = 1:8, size = nrow(new.pred), replace = T)
  pred <- rbind(pred, new.pred)
  # Remove dead prey
  inds <- inds[inds[,dcol] == 0,]
  
  pred.prey <- list(pred = pred, inds = inds)
    return(pred.prey)  # To call upon these seperatly one can use pred.prey$pred and pred.prey$inds 
}


```

#Testing model
Now we can test this model to see if it works as intended by adding some prey and predator so same place

```{r}

inds           <- array(data = 0, dim = c(6, 5));
colnames(inds) <- c("body_mass", "x_loc", "y_loc", "repr", "death");
inds[,1]       <- rnorm(n = dim(inds)[1], mean = 23, sd = 3);
inds[1:3, 2]   <- 1 # x location
inds[1:3, 3]   <- 1 # y location
inds[4:5, 2]   <- 2 
inds[4:5, 3]   <- 1 
inds[6, 2]     <- 3 
inds[6, 3]     <- 1

pred           <- array(data = 0, dim = c(4, 5));
colnames(pred) <- c("body_mass", "x_loc", "y_loc", "repr", "death");
pred[,1]       <- rnorm(n = dim(pred)[1], mean = 20, sd = 2.5);
pred[1, 2]     <- 1 # x location
pred[1, 3]     <- 1 # y location
pred[2, 2]     <- 2
pred[2, 3]     <- 1
pred[3, 2]     <- 3
pred[3, 3]     <- 1
pred[4, 2]     <- 4
pred[4, 3]     <- 1

pred.prey <- predation(inds = inds, pred = pred)
inds      <- pred.prey$inds
pred      <- pred.prey$pred

print(inds)
print(pred)

```

With that we should now be able to -
##Put it all together \o/ 

```{r}

# ------------------- Initialise indivduals (prey)
inds <-  matrix(data = 0, nrow = 40, ncol = 5)
colnames(inds) <- c("bodymass", "x location", "y location", "Birth", "Death")
rownames(inds) <- paste0("Prey ", 1:nrow(inds))

inds[, 1] <- rnorm(n = nrow(inds), mean = 23, sd = 3)  #This approximates a roe deer ish
inds[, 2] <- sample(x = 1:8, size = nrow(inds), replace = T)
inds[, 3] <- sample(x = 1:8, size = nrow(inds), replace = T)

# ------------------ Initialise indivduals (predator)
pred <-  matrix(data = 0, nrow = 20, ncol = 5)
colnames(pred) <- c("bodymass", "x location", "y location", "Birth", "Death")
rownames(pred) <- paste0("Predator ", 1:nrow(pred))

pred[, 1] <- rnorm(n = nrow(pred), mean = 20, sd = 2.5)  #This approximates a lynx ish
pred[, 2] <- sample(x = 1:8, size = nrow(pred), replace = T)
pred[, 3] <- sample(x = 1:8, size = nrow(pred), replace = T)

# ------------------- Start complete simulation 

time.steps <- 40
inds.hist <- NULL
pred.hist <- NULL

for (i in 1:time.steps){
  pred <- movement(inds = pred)
  inds <- movement(inds = inds)
  inds <- birth(inds = inds, lambda = 1.5)
  pred.prey <- predation(pred = pred, inds = inds)
  pred <- pred.prey$pred
  inds <- pred.prey$inds
  inds <- death(inds)
  inds <- inds[inds[, 5] == 0,]  #retain only living
  
  inds.hist[[i]] <- inds
  pred.hist[[i]] <- pred
}

# Plotting it

abundance <- matrix(data = NA, nrow = time.steps, ncol = 3)
colnames(abundance) <- c("Year", "Prey", "Predator")

for (i in 1:time.steps){
  abundance[i,1] <- i
  abundance[i,2] <- nrow(inds.hist[[i]])
  abundance[i,3] <- nrow(pred.hist[[i]])
}

{
plot(x = abundance[,1], y = abundance[, 2], type = "l",
     xlab = "Time", ylab = "Abundance", 
     col = "mediumpurple4", ylim = c(0, 70))
lines(x = abundance[,1], y = abundance[, 3], col = "hotpink2")
legend(x = 5, y = 70, legend = c("Prey", "Predator"), 
       col = c("mediumpurple4", "hotpink2"), 
       lty = c(1, 1), box.lty = 0)
}


```





