---
title: "IBMs"
author: "Bell the best"
date: "2023-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Individual based models eg Agent-based models

Models systems of discrete individuals and simulate changes in indivudal composition and properties. Can be used in conjunction with experimental data to predict dynamics of complex systems. 

*Perks*:(1) Includes among individual variation x. genotype, phenotype or location. 
        (2)Indivduals are discrete, so stochasticty is naturally apart of the model. x. genetic drift.
        
Individuals are represented in a list, table, matrix etc. 

### For example:

```{r}
inds <- array(data = 0, dim =c (5,3))
colnames(inds) <- paste0("Characteristic ", 1:ncol(inds))
rownames(inds) <- paste0("Individual ", 1:nrow(inds))

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
